---
title: "QTE benchmark: py-qte vs R qte"
engine: knitr
params:
  n_iter: 10
  n_iter_simple: 1000
  n_workers: 4
  seed: 42
  n_sim_unconfounded: 2000
  n_sim_confounded: 2000
  n_iter_sim_unconfounded: 3
  n_iter_sim_confounded: 3
  n_iter_sim_py_qte_only: 3
  n_sim_py_qte_only: 0
format:
  html:
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
---

{{< include _setup.qmd >}}

```{r}
#| label: load-qte
#| code-summary: "Show code (R loading the qte package)"
library(qte)
```

```{r}
#| label: set-blas-open-mp-env-vars
#| code-summary: "Show code (R configuring single-threaded BLAS/OpenMP)"
# For parallel bootstrap in qte (r-package) we need BLAS/OpenMP to be single threaed.
library(RhpcBLASctl)
blas_set_num_threads(1L)
omp_set_num_threads(1L)
pbapply::pboptions(type = "none")
```

```{r}
#| label: parameters
#| code-summary: "Show code (R reading the benchmark parameters)"
N_ITER <- as.integer(params$n_iter)
N_ITER_SIMPLE <- as.integer(params$n_iter_simple)
N_WORKERS <- as.integer(params$n_workers)
SEED <- as.integer(params$seed)
N_SIM_UNCONFOUNDED <- as.integer(params$n_sim_unconfounded)
N_SIM_CONFOUNDED <- as.integer(params$n_sim_confounded)
N_SIM_PY_QTE_ONLY <- as.integer(params$n_sim_py_qte_only)
N_ITER_SIM_UNCONFOUNDED <- as.integer(params$n_iter_sim_unconfounded)
N_ITER_SIM_CONFOUNDED <- as.integer(params$n_iter_sim_confounded)
N_ITER_SIM_PY_QTE_ONLY <- as.integer(params$n_iter_sim_py_qte_only)
```

```{r}
#| code-summary: "Show code (R quantiles and formulas)"
QS <- c(0.25, 0.5, 0.75)
QS_SIMPLE <- c(0.25, 0.5, 0.75, 0.9)
X_R <- ~ age + I(age^2) + education + black + hispanic + married + nodegree
X_PY <- "age + I(age**2) + education + black + hispanic + married + nodegree"
```

```{python}
#| label: py-setup
#| code-summary: "Show code (Python importing libraries and loading the data)"
import time

import polars as pl
from great_tables import GT, loc, md, px, style

from qte.bootstrap import BootstrapConfig
from qte.cross_sectional import estimate_aipw_qte, estimate_simple_qte
from qte.datasets import load_lalonde

DATA = load_lalonde(controls_source="psid")
qs = [float(q) for q in r.QS]
qs_simple = [float(q) for q in r.QS_SIMPLE]
```

```{python}
#| label: py-helpers
#| code-summary: "Show code (Python defining the comparison-table helpers)"
def make_source_note(n_obs: int, n_iter: int, n_workers: int) -> str:
    return (
        f"N = {int(n_obs):,}; {int(n_iter):,} bootstrap iterations; "
        f"{int(n_workers)} workers for parallel computations."
    )


def make_timing_table(ds: pl.DataFrame, estimation: str, source_note: str) -> GT:
    return (
        GT(ds.select("package", "computation", "seconds"))
        .cols_label(package="Package", computation="Computation", seconds="Seconds")
        .fmt_number(columns=["seconds"], decimals=1, use_seps=True)
        .tab_header(title=md("**Speed Comparison**"), subtitle=estimation)
        .tab_source_note(source_note)
    )


def make_comparison_data(qs, series) -> pl.DataFrame:
    """Tidy estimate/SE data for a package-vs-package comparison.

    ``series`` maps a ``(package, computation)`` label to an ``(effects, ses)``
    pair, each ordered as ``[QTE(q) for q in qs] + [ATT]``.
    """
    n_qte = len(qs)
    records = []
    for (package, computation), (effects, ses) in series.items():
        for i, q in enumerate(qs):  # quantiles
            records.append(
                {
                    "effect": "QTE",
                    "quantile": f"{q:.2f}",
                    "package": package,
                    "computation": computation,
                    "estimate": float(effects[i]),
                    "se": float(ses[i]),
                }
            )
        records.append(
            {
                "effect": "ATT",
                "quantile": "",
                "package": package,
                "computation": computation,
                "estimate": float(effects[n_qte]),
                "se": float(ses[n_qte]),
            }
        )
    return pl.DataFrame(records)


def make_estimate_comparison_table(data: pl.DataFrame, estimation: str, source_note: str) -> GT:
    """Render tidy comparison data (see ``make_comparison_data``).

    Rows are grouped under ``QTE``/``ATT``; package names become the top-level
    column spanners with the computation nested underneath. Estimate and
    standard-error columns get distinct backgrounds, and packages are separated
    by a vertical rule.
    """
    pairs = data.select("package", "computation").unique(maintain_order=True).rows()
    packages = list(dict.fromkeys(package for package, _ in pairs))
    computations = list(dict.fromkeys(computation for _, computation in pairs))

    def series(package, computation):
        # filter preserves the QTE...ATT order produced by make_comparison_data
        return data.filter((pl.col("package") == package) & (pl.col("computation") == computation))

    base = series(*pairs[0])
    frame = {"effect": base["effect"].to_list(), "quantile": base["quantile"].to_list()}
    leaf_labels = {}
    for package, computation in pairs:
        prefix = f"{package}.{computation}"
        sub = series(package, computation)
        frame[f"{prefix}.est"] = sub["estimate"].to_list()
        frame[f"{prefix}.se"] = sub["se"].to_list()
        leaf_labels[f"{prefix}.est"] = "Estimate"
        leaf_labels[f"{prefix}.se"] = md("Standard<br>Error")

    table = pl.DataFrame(frame)
    est_cols = [c for c in table.columns if c.endswith(".est")]
    se_cols = [c for c in table.columns if c.endswith(".se")]

    gt = (
        GT(table, groupname_col="effect")
        .row_group_order(["QTE", "ATT"])
        .cols_label(quantile="Quantile", **leaf_labels)
        .fmt_number(columns=est_cols + se_cols, decimals=1, use_seps=True)
        .tab_header(title=md("**Results Comparison**"), subtitle=estimation)
        .tab_source_note(source_note)
        .tab_style(style=style.fill(color="#E8F1FA"), locations=loc.body(columns=est_cols))
        .tab_style(style=style.fill(color="#FBF0E1"), locations=loc.body(columns=se_cols))
        .tab_style(style=style.fill(color="#D3E4F5"), locations=loc.column_labels(columns=est_cols))
        .tab_style(style=style.fill(color="#F6E0C4"), locations=loc.column_labels(columns=se_cols))
    )
    for package in packages:
        # narrower spanners first so the package spanner sits on top
        package_columns = []
        for computation in computations:
            est_col, se_col = f"{package}.{computation}.est", f"{package}.{computation}.se"
            if est_col not in frame:
                continue
            gt = gt.tab_spanner(
                computation, columns=[est_col, se_col], id=f"{package}|{computation}"
            )
            package_columns += [est_col, se_col]
        gt = gt.tab_spanner(package, columns=package_columns, id=f"{package}|pkg")
    if len(packages) > 1:
        divider = "#8A8A8A"
        second_package = packages[1]
        first_computation = next(
            computation for package, computation in pairs if package == second_package
        )
        first_of_second = f"{second_package}.{first_computation}.est"
        gt = gt.tab_style(
            style=style.borders(sides="left", color=divider, style="solid", weight=px(2)),
            locations=[
                loc.body(columns=first_of_second),
                loc.column_labels(columns=first_of_second),
                loc.spanner_labels(ids=[f"{second_package}|pkg"]),
                loc.spanner_labels(ids=[f"{second_package}|{first_computation}"]),
            ],
        )
    return gt
```

## Summary

Here, we compare the performance of the `py-qte` and the `qte` package. We use a medium-sized
machine (see the [Environment](#environment) section for details). The comparisons are based on the
relatively small Lalonde dataset that comes with the `qte` package and on a medium-sized simulated
dataset inspired by [this](https://docs.doubleml.org/stable/examples/py_double_ml_pq.html#Data)
section from the documentation of the DoubleML package.

Notice that in particular estimators that involve an outcome model are relatively computationally
intensive since they involve fitting a quantile regression model on a grid of quantiles in each
bootstrap run.

Even for the _small Lalonde_ dataset (`{r} formatC(py$DATA$height, format = "d", big.mark = ",")`
observations), we find sizeable differences between the `qte` and the `py-qte` package in sequential
computations. These can be tightly connected to different implementation approaches. Parallelized
approaches benefit `qte` relatively more (in this setting). See @tbl-simple-time. For estimation
with confounders, the same holds true, but even after parallelization the gap remains sizeable, and
the `qte` package takes about 3 times as long as the `py-qte` package (see @tbl-aipw-timings).

For a larger simulated dataset (`{r} formatC(N_SIM_UNCONFOUNDED, format = "d", big.mark = ",")`
observations), we find that without confounders `py-qte` is about X times as fast even after
parallelization. See @tbl-sim-simple-time. For the AIPW estimator we find that the parallelized
`py-qte` version on a medium-sized dataset
(`{r} formatC(N_SIM_CONFOUNDED, format = "d", big.mark = ",")`) with a reasonable number of
bootstrap iterations is more than ten times faster than the `qte` version. See @tbl-sim-cov-time.

Lastly, the Python version can provide results for
`{r} formatC(N_SIM_PY_QTE_ONLY, format = "d", big.mark = ",")` observations in about the same time
that the parallelized `qte` version took for
`{r} formatC(N_SIM_CONFOUNDED, format = "d", big.mark = ",")` observations. See @tbl-py-only-time.

These results stress the importance of algorithmic choices and also parallelization strategies for
the bootstrap iterations.

## Cross-Sectional Data

```{r}
#| include: false
num_rows_lalonde <- py$DATA$height
```

We compare cross-sectional estimators on the Lalonde dataset. The Lalonde dataset is a relatively
small dataset with only `{r} formatC(num_rows_lalonde, format = "d", big.mark = ",")` observations.

### Estimation under Unconfoundedness

```{r r-simple}
#| code-summary: "Show code (R estimate of the unconditional QTE)"
#| time_it: true
# Unconditional QTE with no covariates == R's IPW branch with x = NULL.
run_r_simple <- function(cl) {
  set.seed(SEED)
  t0 <- Sys.time()
  res <- unc_qte(
    yname = "re78", dname = "treat", data = lalonde.psid, xformla = NULL,
    probs = QS_SIMPLE, est_method = "ipw", target = "qtt",
    cl = cl, biters = N_ITER_SIMPLE
  )
  list(
    seconds = as.numeric(Sys.time() - t0, units = "secs"),
    vals = list(
      qte = as.numeric(res$qte), se = as.numeric(res$qte.se),
      ate = as.numeric(res$ate), ate_se = as.numeric(res$ate.se)
    )
  )
}

r_simple_seq <- run_r_simple(1)
r_simple_par <- run_r_simple(N_WORKERS)
r_simple_seq_seconds <- r_simple_seq$seconds
r_simple_par_seconds <- r_simple_par$seconds
r_simple_seq_vals <- r_simple_seq$vals
r_simple_par_vals <- r_simple_par$vals
```

```{python}
#| label: py-simple
#| code-summary: "Show code (Python estimate of the unconditional QTE)"
#| time_it: true
def run_py_simple(n_workers):
    cfg = BootstrapConfig(n_iter=int(r.N_ITER_SIMPLE), seed=int(r.SEED), n_workers=n_workers)
    t0 = time.perf_counter()
    res = estimate_simple_qte(DATA, "re78", "treat", qs=qs_simple, bootstrap_config=cfg)
    return time.perf_counter() - t0, res


py_simple_seq_seconds, res_py_simple_seq = run_py_simple(1)
py_simple_par_seconds, res_py_simple_par = run_py_simple(int(r.N_WORKERS))
```

```{python}
#| label: simple-tables
#| code-summary: "Show code (Python building the simple comparison tables)"
simple_order = [f"QTE({q:.2f})" for q in qs_simple] + ["ATE"]


def py_simple_cols(res):
    rows = {
        f"QTE({row['q']:.2f})": (row["effect"], row["se"]) for row in res.qtt.iter_rows(named=True)
    }
    rows["ATE"] = (res.att["effect"][0], res.att["se"][0])
    return [rows[k][0] for k in simple_order], [rows[k][1] for k in simple_order]


def r_simple_cols(vals):
    eff = [float(x) for x in vals["qte"]] + [float(vals["ate"])]
    err = [float(x) for x in vals["se"]] + [float(vals["ate_se"])]
    return eff, err


simple_source_note = make_source_note(DATA.height, int(r.N_ITER_SIMPLE), int(r.N_WORKERS))
simple_table_html = make_estimate_comparison_table(
    make_comparison_data(
        qs_simple,
        {
            ("py-qte", "Sequential"): py_simple_cols(res_py_simple_seq),
            ("py-qte", "Parallel"): py_simple_cols(res_py_simple_par),
            ("qte", "Sequential"): r_simple_cols(r.r_simple_seq_vals),
            ("qte", "Parallel"): r_simple_cols(r.r_simple_par_vals),
        },
    ),
    "Estimation under Unconfoundedness",
    simple_source_note,
).as_raw_html()

data = pl.DataFrame(
    {
        "language": ["Python", "Python", "R", "R"],
        "package": ["py-qte", "py-qte", "qte", "qte"],
        "computation": ["Sequential", "Parallel", "Sequential", "Parallel"],
        "workers": [1, int(r.N_WORKERS), 1, int(r.N_WORKERS)],
        "iterations": [int(r.N_ITER_SIMPLE)] * 4,
        "seconds": [
            py_simple_seq_seconds,
            py_simple_par_seconds,
            float(r.r_simple_seq_seconds),
            float(r.r_simple_par_seconds),
        ],
    }
)

simple_timing_html = make_timing_table(
    data, "Estimation under Unconfoundedness", simple_source_note
).as_raw_html()
```

```{r}
#| label: tbl-simple-res
#| tbl-cap: "Unconditional QTE Estimates on the Lalonde Data."
#| results: asis
#| echo: false
cat(py$simple_table_html)
```

```{r}
#| label: tbl-simple-time
#| tbl-cap: "Unconditional QTE Runtimes on the Lalonde Data."
#| results: asis
#| echo: false
cat(py$simple_timing_html)
```

Notice first that `py-qte` replicates the point estimates of the `qte` package and produces similar
standard errors. Notice that the `py-qte` produces consistent standard error estimates across the
sequential and the parallelized run as it should.

We see that `py-qte` is in a _sequential scenario_ much faster than `qte`. This is probably mostly
because `np.quantile` provides an efficient implementation of computing weighted quantiles whereas
the `qte` package runs a solver for each quantile. The speed-up from parallelization is very large
for `qte` and sizeable but relatively low `py-qte`.

### AIPW Estimator

```{r r-sequential}
#| code-summary: "Show code (R AIPW estimate (sequential))"
#| time_it: true
set.seed(SEED)
t0 <- Sys.time()
res_r_seq <- unc_qte(
  yname = "re78", dname = "treat", data = lalonde.psid, xformla = X_R,
  probs = QS, est_method = "aipw", target = "qte", cl = 1, biters = N_ITER
)
r_seq_seconds <- as.numeric(Sys.time() - t0, units = "secs")

r_seq_vals <- list(
  qte = as.numeric(res_r_seq$qte), se = as.numeric(res_r_seq$qte.se),
  ate = as.numeric(res_r_seq$ate), ate_se = as.numeric(res_r_seq$ate.se)
)
```

```{r r-parallel}
#| code-summary: "Show code (R AIPW estimate (parallel))"
#| time_it: true
set.seed(SEED)
t0 <- Sys.time()
res_r_par <- unc_qte(
  yname = "re78", dname = "treat", data = lalonde.psid, xformla = X_R,
  probs = QS, est_method = "aipw", target = "qte", cl = N_WORKERS, biters = N_ITER
)
r_par_seconds <- as.numeric(Sys.time() - t0, units = "secs")

r_par_vals <- list(
  qte = as.numeric(res_r_par$qte), se = as.numeric(res_r_par$qte.se),
  ate = as.numeric(res_r_par$ate), ate_se = as.numeric(res_r_par$ate.se)
)
```

```{python}
#| label: py-sequential
#| code-summary: "Show code (Python AIPW estimate (sequential))"
#| time_it: true
cfg = BootstrapConfig(n_iter=int(r.N_ITER), seed=int(r.SEED), n_workers=1)
t0 = time.perf_counter()
res_py_seq = estimate_aipw_qte(
    DATA,
    "re78",
    "treat",
    qs=qs,
    ps_x_formular=r.X_PY,
    or_x_formular=r.X_PY,
    bootstrap_config=cfg,
)
py_seq_seconds = time.perf_counter() - t0
```

```{python}
#| label: py-parallel
#| code-summary: "Show code (Python AIPW estimate (parallel))"
#| time_it: true
cfg = BootstrapConfig(n_iter=int(r.N_ITER), seed=int(r.SEED), n_workers=int(r.N_WORKERS))
t0 = time.perf_counter()
res_py_par = estimate_aipw_qte(
    DATA,
    "re78",
    "treat",
    qs=qs,
    ps_x_formular=r.X_PY,
    or_x_formular=r.X_PY,
    bootstrap_config=cfg,
)
py_par_seconds = time.perf_counter() - t0
```

```{python}
#| label: aipw-tables
#| code-summary: "Show code (Python building the AIPW comparison tables)"
order = [f"QTE({q:.2f})" for q in qs] + ["ATE"]


def py_cols(res):
    rows = {
        f"QTE({row['q']:.2f})": (row["effect"], row["se"]) for row in res.qtt.iter_rows(named=True)
    }
    rows["ATE"] = (res.att["effect"][0], res.att["se"][0])
    return [rows[k][0] for k in order], [rows[k][1] for k in order]


def r_cols(vals):
    eff = [float(x) for x in vals["qte"]] + [float(vals["ate"])]
    err = [float(x) for x in vals["se"]] + [float(vals["ate_se"])]
    return eff, err


py_seq_e, py_seq_s = py_cols(res_py_seq)
py_par_e, py_par_s = py_cols(res_py_par)
r_seq_e, r_seq_s = r_cols(r.r_seq_vals)
r_par_e, r_par_s = r_cols(r.r_par_vals)

aipw_data = make_comparison_data(
    qs,
    {
        ("py-qte", "Sequential"): (py_seq_e, py_seq_s),
        ("py-qte", "Parallel"): (py_par_e, py_par_s),
        ("qte", "Sequential"): (r_seq_e, r_seq_s),
        ("qte", "Parallel"): (r_par_e, r_par_s),
    },
)

aipw_source_note = make_source_note(DATA.height, int(r.N_ITER), int(r.N_WORKERS))
table_html = make_estimate_comparison_table(
    aipw_data, "Estimation via AIPW", aipw_source_note
).as_raw_html()

timing = pl.DataFrame(
    {
        "language": ["Python", "Python", "R", "R"],
        "package": ["py-qte", "py-qte", "qte", "qte"],
        "computation": ["Sequential", "Parallel", "Sequential", "Parallel"],
        "workers": [1, int(r.N_WORKERS), 1, int(r.N_WORKERS)],
        "iterations": [int(r.N_ITER)] * 4,
        "seconds": [py_seq_seconds, py_par_seconds, float(r.r_seq_seconds), float(r.r_par_seconds)],
    }
)

timing_html = make_timing_table(timing, "Estimation via AIPW", aipw_source_note).as_raw_html()
```

```{r}
#| label: tbl-aipw-res
#| tbl-cap: "AIPW Comparison of `py-qte` and `qte`."
#| results: asis
#| echo: false
cat(py$table_html)
```

In terms of results, we see in @tbl-aipw-res that the point estimates are close between `py-qte` and
`qte`. The standard errors are once more consistent across the sequential and parallelized runs.

```{r results='asis', echo=FALSE}
#| label: tbl-aipw-timings
#| tbl-cap: "AIPW Comparison of `py-qte` and `qte`."
#| results: asis
cat(py$timing_html)
```

We further see in @tbl-aipw-timings that the sequential version of `py-qte` is much faster than the
`qte` version. This is probably largely due to the fact that `py-qte` computes $F_{AIPW}(y)$ on the
y-grid in a single (vectorized) pass whereas the `qte` version computes each point on the y-grid
separately in a loop. The parallelized version of `py-qte` retains a sizeable advantage over the
`qte` version although the gap is sizeably narrower than in the sequential case.

## Simulated Data

The sections above use the Lalonde dataset, which has only a few thousand observations. To see how
well the `py-qte` and the `qte` package do on large dataset we resort to simulated data.

```{python}
#| label: sim-data
#| code-summary: "Show code (Python simulating the datasets)"
#| time_it: true
from qte.cross_sectional import simulate_covariate_data, simulate_simple_data


def add_correct_features(ds):
    return ds.with_columns(
        ind1=(pl.col("x1") > 0.1).cast(pl.Float64),
        ind2=((pl.col("x0") * pl.col("x2")) > 0).cast(pl.Float64),
        sq=(2.5 + 0.3 * pl.col("x1")).sqrt(),
    )


EFFECT_SCALE = 1000.0
sim_simple = simulate_simple_data(
    int(r.N_SIM_UNCONFOUNDED), treatment_effect=EFFECT_SCALE, seed=int(r.SEED)
)
sim_cov = add_correct_features(
    simulate_covariate_data(
        int(r.N_SIM_CONFOUNDED), effect_scale=EFFECT_SCALE, seed=int(r.SEED) + 1
    )
)

# Plain dicts so reticulate hands R a data.frame (polars objects are not converted).
SIM_SIMPLE_R = sim_simple.to_dict(as_series=False)
SIM_COV_R = sim_cov.to_dict(as_series=False)
```

```{r r-sim}
#| code-summary: "Show code (R estimates on the simulated datasets)"
#| time_it: true
sim_simple_df <- as.data.frame(py$SIM_SIMPLE_R)
sim_cov_df <- as.data.frame(py$SIM_COV_R)

run_r_sim <- function(data, xformla, est_method, target, cl, biters) {
  set.seed(SEED)
  t0 <- Sys.time()
  res <- unc_qte(
    yname = "y", dname = "treat", data = data, xformla = xformla,
    probs = QS, est_method = est_method, target = target,
    cl = cl, biters = biters
  )
  list(
    seconds = as.numeric(Sys.time() - t0, units = "secs"),
    vals = list(
      qte = as.numeric(res$qte), se = as.numeric(res$qte.se),
      ate = as.numeric(res$ate), ate_se = as.numeric(res$ate.se)
    )
  )
}

r_sim_simple_par <- run_r_sim(sim_simple_df, NULL, "ipw", "qtt", N_WORKERS, N_ITER_SIM_UNCONFOUNDED)
r_sim_cov_par <- run_r_sim(sim_cov_df, ~ x3 + x4 + ind1 + ind2 + sq, "aipw", "qte", N_WORKERS, N_ITER_SIM_CONFOUNDED)
```

```{python}
#| label: py-sim
#| code-summary: "Show code (Python estimates on the simulated datasets)"
#| time_it: true
def run_py_sim_simple(n_workers):
    cfg = BootstrapConfig(
        n_iter=int(r.N_ITER_SIM_UNCONFOUNDED), seed=int(r.SEED), n_workers=n_workers
    )
    t0 = time.perf_counter()
    res = estimate_simple_qte(sim_simple, "y", "treat", qs=qs, bootstrap_config=cfg)
    return time.perf_counter() - t0, res


def run_py_sim_cov(n_workers):
    cfg = BootstrapConfig(
        n_iter=int(r.N_ITER_SIM_CONFOUNDED), seed=int(r.SEED), n_workers=n_workers
    )
    t0 = time.perf_counter()
    res = estimate_aipw_qte(
        sim_cov,
        "y",
        "treat",
        qs=qs,
        ps_x_formular="x1 + x3 + I(x0 > 0)",
        or_x_formular="x3 + x4 + ind1 + ind2 + sq",
        bootstrap_config=cfg,
    )
    return time.perf_counter() - t0, res


py_sim_simple_seq_seconds, res_py_sim_simple_seq = run_py_sim_simple(1)
py_sim_simple_par_seconds, res_py_sim_simple_par = run_py_sim_simple(int(r.N_WORKERS))
py_sim_cov_seq_seconds, res_py_sim_cov_seq = run_py_sim_cov(1)
py_sim_cov_par_seconds, res_py_sim_cov_par = run_py_sim_cov(int(r.N_WORKERS))
```

```{python}
#| label: sim-tables
#| code-summary: "Show code (Python building the simulated-data comparison tables)"
def make_sim_tables(title, py_seq, py_par, r_par, source_note):
    py_seq_seconds, py_seq_res = py_seq
    py_par_seconds, py_par_res = py_par
    series = {
        ("py-qte", "Sequential"): py_cols(py_seq_res),
        ("py-qte", "Parallel"): py_cols(py_par_res),
    }
    rows = {
        "language": ["Python", "Python"],
        "package": ["py-qte", "py-qte"],
        "computation": ["Sequential", "Parallel"],
        "workers": [1, int(r.N_WORKERS)],
        "seconds": [py_seq_seconds, py_par_seconds],
    }
    if r_par is not None:
        series[("qte", "Parallel")] = r_cols(r_par["vals"])
        rows["language"].append("R")
        rows["package"].append("qte")
        rows["computation"].append("Parallel")
        rows["workers"].append(int(r.N_WORKERS))
        rows["seconds"].append(float(r_par["seconds"]))
    table_html = make_estimate_comparison_table(
        make_comparison_data(qs, series), title, source_note
    ).as_raw_html()
    timing_html = make_timing_table(pl.DataFrame(rows), title, source_note).as_raw_html()
    return table_html, timing_html


sim_simple_table_html, sim_simple_timing_html = make_sim_tables(
    "Simulated data without covariates",
    (py_sim_simple_seq_seconds, res_py_sim_simple_seq),
    (py_sim_simple_par_seconds, res_py_sim_simple_par),
    r.r_sim_simple_par,
    make_source_note(r.N_SIM_UNCONFOUNDED, r.N_ITER_SIM_UNCONFOUNDED, r.N_WORKERS),
)
sim_cov_table_html, sim_cov_timing_html = make_sim_tables(
    "Simulated data with covariates",
    (py_sim_cov_seq_seconds, res_py_sim_cov_seq),
    (py_sim_cov_par_seconds, res_py_sim_cov_par),
    r.r_sim_cov_par,
    make_source_note(r.N_SIM_CONFOUNDED, r.N_ITER_SIM_CONFOUNDED, r.N_WORKERS),
)
```

### Estimation under Unconfoundedness

```{r}
#| label: tbl-sim-simple-res
#| tbl-cap: "QTE Estimates on the Simulated Unconfounded Data."
#| results: asis
#| echo: false
cat(py$sim_simple_table_html)
```

```{r}
#| label: tbl-sim-simple-time
#| tbl-cap: "QTE Runtimes on the Simulated Unconfounded Data."
#| results: asis
#| echo: false
cat(py$sim_simple_timing_html)
```

As in the Lalonde comparison, the two packages agree on the point estimates and standard errors. We
omit the sequential `qte` run here because it is very slow on larger data; even against the parallel
`qte` run `py-qte` is faster. Because the simulated design is randomised, the treatment effect is
identified by the unadjusted comparison; `simulate_simple_data` generates it with the treatment
effect set by `EFFECT_SCALE`.

### AIPW Estimator

```{python}
#| label: sim-true
#| code-summary: "Show code (Python true effects of the confounded simulation)"
#| time_it: true
import gc

import numpy as np

N_TRUE = int(10e6)
true_ds = simulate_covariate_data(N_TRUE, effect_scale=EFFECT_SCALE, seed=int(r.SEED))
true_y0 = true_ds["y_0"].to_numpy()
true_y1 = true_ds["y_1"].to_numpy()
true_ate = float(np.mean(true_y1 - true_y0))
true_qte = np.quantile(true_y1, qs) - np.quantile(true_y0, qs)

# Drop the large simulated population before the later runs.
del true_ds, true_y0, true_y1
gc.collect()

true_rows = pl.concat(
    [
        pl.DataFrame({"Quantity": ["ATE"], "Estimate": [true_ate]}),
        pl.DataFrame({"Quantity": [f"QTE({q:.2f})" for q in qs], "Estimate": true_qte}),
    ]
)
true_table_html = (
    GT(true_rows)
    .fmt_number(columns=["Estimate"], decimals=1, use_seps=True)
    .tab_header(
        title=md("**True Effects**"),
        subtitle="Confounded simulation (population QTE)",
    )
    .tab_source_note(f"True values from a simulation with N = {N_TRUE:,} units.")
    .as_raw_html()
)
```

```{r}
#| label: tbl-sim-true
#| tbl-cap: "True Mean and Quantile Treatment Effects of the Confounded Simulation."
#| results: asis
#| echo: false
cat(py$true_table_html)
```

```{r}
#| label: tbl-sim-cov-res
#| tbl-cap: "AIPW QTE Estimates on the Simulated Confounded Data."
#| results: asis
#| echo: false
cat(py$sim_cov_table_html)
```

```{r}
#| label: tbl-sim-cov-time
#| tbl-cap: "AIPW QTE Runtimes on the Simulated Confounded Data."
#| results: asis
#| echo: false
cat(py$sim_cov_timing_html)
```

The covariate design is confounded and the treated outcome is more dispersed, so the quantile
treatment effect increases with the quantile while the average effect is
`{r} py$EFFECT_SCALE * 0.5`. Here too `py-qte` is faster than the parallel `qte` run.

```{python}
#| label: py-only-sim
#| code-summary: "Show code (Python-only simulated-data run)"
#| time_it: true
if int(r.N_SIM_PY_QTE_ONLY) > 0:
    sim_py_only = add_correct_features(
        simulate_covariate_data(
            int(r.N_SIM_PY_QTE_ONLY), effect_scale=EFFECT_SCALE, seed=int(r.SEED) + 2
        )
    )
    sim_cfg = BootstrapConfig(
        n_iter=int(r.N_ITER_SIM_PY_QTE_ONLY), seed=int(r.SEED), n_workers=int(r.N_WORKERS)
    )
    t0 = time.perf_counter()
    res_py_only_seq = estimate_aipw_qte(
        sim_py_only,
        "y",
        "treat",
        qs=qs,
        ps_x_formular="x1 + x3 + I(x0 > 0)",
        or_x_formular="x3 + x4 + ind1 + ind2 + sq",
        bootstrap_config=sim_cfg._replace(n_workers=1),
    )
    py_only_seq_seconds = time.perf_counter() - t0
    t0 = time.perf_counter()
    res_py_only_par = estimate_aipw_qte(
        sim_py_only,
        "y",
        "treat",
        qs=qs,
        ps_x_formular="x1 + x3 + I(x0 > 0)",
        or_x_formular="x3 + x4 + ind1 + ind2 + sq",
        bootstrap_config=sim_cfg,
    )
    py_only_par_seconds = time.perf_counter() - t0
    py_only_table_html, py_only_timing_html = make_sim_tables(
        "py-qte only (no R)",
        (py_only_seq_seconds, res_py_only_seq),
        (py_only_par_seconds, res_py_only_par),
        None,
        make_source_note(r.N_SIM_PY_QTE_ONLY, r.N_ITER_SIM_PY_QTE_ONLY, r.N_WORKERS),
    )
else:
    py_only_table_html = ""
    py_only_timing_html = ""
```

```{r results='asis', echo=FALSE}
if (N_SIM_PY_QTE_ONLY > 0) {
  cat("### py-qte only\n\n")
  cat("With `n_sim_py_qte_only` set we run a separate confounded simulation of that size with only `py-qte`. The tables therefore contain only `py-qte` columns.\n\n")
}
```

```{r}
#| label: tbl-py-only-res
#| tbl-cap: "AIPW QTE Estimates for the py-qte-only Run."
#| results: asis
#| echo: false
if (N_SIM_PY_QTE_ONLY > 0) cat(py$py_only_table_html)
```

```{r}
#| label: tbl-py-only-time
#| tbl-cap: "AIPW QTE Runtimes for the py-qte-only Run."
#| results: asis
#| echo: false
if (N_SIM_PY_QTE_ONLY > 0) cat(py$py_only_timing_html)
```

## Environment

```{r}
#| label: environment
#| code-summary: "Show code (R environment details)"
system_ <- Sys.info()[["sysname"]]


get_linux_os <- function(key) {
  if (!file.exists("/etc/os-release")) {
    return(NA_character_)
  }
  x <- grep(paste0("^", key, "="), readLines("/etc/os-release"), value = TRUE)[1]
  if (is.na(x)) {
    return(NA_character_)
  }
  value <- gsub('^"|"$', "", sub(paste0("^", "PRETTY_NAME", "="), "", x))
  if (grepl("Void Linux", value)) value <- paste0(value, " btw.")
  value
}
proc_value <- function(path, key) {
  if (!file.exists(path)) {
    return(NA_character_)
  }
  x <- grep(key, readLines(path), value = TRUE)[1]
  if (is.na(x)) {
    return(NA_character_)
  }
  trimws(sub("^[^:]*:", "", x))
}

if (system_ == "Darwin") {
  cpu_name <- system("sysctl -n machdep.cpu.brand_string", intern = TRUE)
} else {
  cpu_name <- proc_value("/proc/cpuinfo", "model name")
}

library(ps)
env_stats <- data.frame(
  item = c(
    "OS", "Kernel", "CPU", "Cores (physical)", "Cores (logical)",
    "Memory", "R", "BLAS", "Python"
  ),
  value = c(
    if (system_ == "Linux") get_linux_os("PRETTY_NAME") else system_,
    paste(Sys.info()[["sysname"]], Sys.info()[["release"]], Sys.info()[["machine"]]),
    cpu_name,
    parallel::detectCores(logical = FALSE),
    parallel::detectCores(logical = TRUE),
    sprintf("%.1f GiB", ps_system_memory()$total / 1e9),
    R.version.string,
    unname(extSoftVersion()["BLAS"]),
    sub(" .*", "", reticulate::py_config()$version_string)
  )
)
knitr::kable(env_stats, col.names = c("", ""), row.names = FALSE)
```
