---
title: "QTE benchmark: py-qte vs R qte"
engine: knitr
params:
  n_iter: 10
  n_iter_simple: 1000
  n_workers: 4
  seed: 42
format:
  html:
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
# Render with `quarto render docs`. This document requires two prepared
# environments and fails loudly if either is missing:
#   - Python: `uv sync` (creates ../.venv)
#   - R:      `renv::restore()` inside docs/ (installs the pinned packages)
# Override params from the CLI, e.g.:
#   quarto render docs -P n_iter:100 -P n_iter_simple:1000 -P n_workers:8
docs_dir <- dirname(knitr::current_input(dir = TRUE))

# Force all libraries to be loaded from the renv environment.
options(renv.config.synchronized.check = FALSE)
renv::load(project = docs_dir, quiet = TRUE)

# Get the Python environment it is currently located once step above the docs directory
library(reticulate)
venv <- file.path(docs_dir, "..", ".venv")
use_virtualenv(normalizePath(venv), required = TRUE)
```

```{r}
#| label: load-qte
#| code-summary: "Show code (R loading the qte package)"
library(qte)
```

```{r}
#| label: set-blas-open-mp-env-vars
#| code-summary: "Show code (R configuring single-threaded BLAS/OpenMP)"
# For parallel bootstrap in qte (r-package) we need BLAS/OpenMP to be single threaed.
library(RhpcBLASctl)
blas_set_num_threads(1L)
omp_set_num_threads(1L)
pbapply::pboptions(type = "none")
```

```{r}
#| label: parameters
#| code-summary: "Show code (R reading the benchmark parameters)"
N_ITER <- as.integer(params$n_iter)
N_ITER_SIMPLE <- as.integer(params$n_iter_simple)
N_WORKERS <- as.integer(params$n_workers)
SEED <- as.integer(params$seed)
```

```{r}
#| code-summary: "Show code (R quantiles and formulas)"
QS <- c(0.25, 0.5, 0.75)
QS_SIMPLE <- c(0.25, 0.5, 0.75, 0.9)
X_R <- ~ age + I(age^2) + education + black + hispanic + married + nodegree
X_PY <- "age + I(age**2) + education + black + hispanic + married + nodegree"
```

```{python}
#| label: py-setup
#| code-summary: "Show code (Python importing libraries and loading the data)"
import time

import polars as pl
from great_tables import GT, loc, md, px, style

from qte.bootstrap import BootstrapConfig
from qte.cross_sectional import estimate_aipw_qte, estimate_simple_qte
from qte.datasets import load_lalonde

DATA = load_lalonde(controls_source="psid")
qs = [float(q) for q in r.QS]
qs_simple = [float(q) for q in r.QS_SIMPLE]
```

```{python}
#| label: py-helpers
#| code-summary: "Show code (Python defining the comparison-table helpers)"
def make_timing_table(ds: pl.DataFrame, estimation: str) -> GT:
    return (
        GT(ds)
        .cols_label(
            language="Language",
            package="Package",
            computation="Computation",
            workers="Workers",
            iterations="Bootstrap Iterations",
            seconds="Seconds",
        )
        .fmt_number(columns=["seconds"], decimals=1, use_seps=True)
        .fmt_number(columns=["iterations"], decimals=0, use_seps=True)
        .tab_header(title=md("**Speed Comparison**"), subtitle=estimation)
    )


def make_comparison_data(qs, series) -> pl.DataFrame:
    """Tidy estimate/SE data for a package-vs-package comparison.

    ``series`` maps a ``(package, computation)`` label to an ``(effects, ses)``
    pair, each ordered as ``[QTE(q) for q in qs] + [ATT]``.
    """
    n_qte = len(qs)
    records = []
    for (package, computation), (effects, ses) in series.items():
        for i, q in enumerate(qs):  # quantiles
            records.append(
                {
                    "effect": "QTE",
                    "quantile": f"{q:.2f}",
                    "package": package,
                    "computation": computation,
                    "estimate": float(effects[i]),
                    "se": float(ses[i]),
                }
            )
        records.append(
            {
                "effect": "ATT",
                "quantile": "",
                "package": package,
                "computation": computation,
                "estimate": float(effects[n_qte]),
                "se": float(ses[n_qte]),
            }
        )
    return pl.DataFrame(records)


def make_estimate_comparison_table(data: pl.DataFrame, estimation: str) -> GT:
    """Render tidy comparison data (see ``make_comparison_data``).

    Rows are grouped under ``QTE``/``ATT``; package names become the top-level
    column spanners with the computation nested underneath. Estimate and
    standard-error columns get distinct backgrounds, and packages are separated
    by a vertical rule.
    """
    packages = data["package"].unique(maintain_order=True).to_list()
    computations = data["computation"].unique(maintain_order=True).to_list()

    def series(package, computation):
        # filter preserves the QTE...ATT order produced by make_comparison_data
        return data.filter((pl.col("package") == package) & (pl.col("computation") == computation))

    base = series(packages[0], computations[0])
    frame = {"effect": base["effect"].to_list(), "quantile": base["quantile"].to_list()}
    leaf_labels = {}
    for package in packages:
        for computation in computations:
            prefix = f"{package}.{computation}"
            sub = series(package, computation)
            frame[f"{prefix}.est"] = sub["estimate"].to_list()
            frame[f"{prefix}.se"] = sub["se"].to_list()
            leaf_labels[f"{prefix}.est"] = "Estimate"
            leaf_labels[f"{prefix}.se"] = md("Standard<br>Error")

    table = pl.DataFrame(frame)
    est_cols = [c for c in table.columns if c.endswith(".est")]
    se_cols = [c for c in table.columns if c.endswith(".se")]

    gt = (
        GT(table, groupname_col="effect")
        .row_group_order(["QTE", "ATT"])
        .cols_label(quantile="Quantile", **leaf_labels)
        .fmt_number(columns=est_cols + se_cols, decimals=1, use_seps=True)
        .tab_header(title=md("**Results Comparison**"), subtitle=estimation)
        .tab_style(style=style.fill(color="#E8F1FA"), locations=loc.body(columns=est_cols))
        .tab_style(style=style.fill(color="#FBF0E1"), locations=loc.body(columns=se_cols))
        .tab_style(style=style.fill(color="#D3E4F5"), locations=loc.column_labels(columns=est_cols))
        .tab_style(style=style.fill(color="#F6E0C4"), locations=loc.column_labels(columns=se_cols))
    )
    for package in packages:
        # narrower spanners first so the package spanner sits on top
        for computation in computations:
            gt = gt.tab_spanner(
                computation,
                columns=[f"{package}.{computation}.est", f"{package}.{computation}.se"],
                id=f"{package}|{computation}",
            )
        gt = gt.tab_spanner(
            package,
            columns=[
                f"{package}.{computation}.{stat}"
                for computation in computations
                for stat in ("est", "se")
            ],
            id=f"{package}|pkg",
        )
    if len(packages) > 1:
        divider = "#8A8A8A"
        first_of_second = f"{packages[1]}.{computations[0]}.est"
        gt = gt.tab_style(
            style=style.borders(sides="left", color=divider, style="solid", weight=px(2)),
            locations=[
                loc.body(columns=first_of_second),
                loc.column_labels(columns=first_of_second),
                loc.spanner_labels(ids=[f"{packages[1]}|pkg"]),
                loc.spanner_labels(ids=[f"{packages[1]}|{computations[0]}"]),
            ],
        )
    return gt
```

## Cross-Sectional Data

```{r}
#| include: false
num_rows_lalonde <- nrow(py$DATA)
```

We compare cross-sectional estimators on the Lalonde dataset. The Lalonde dataset is a relatively
small dataset with only `{r} num_rows_lalonde` observation.

### Estimation under Unconfoundedness

```{r r-simple}
#| code-summary: "Show code (R estimate of the unconditional QTE)"
# Unconditional QTE with no covariates == R's IPW branch with x = NULL.
run_r_simple <- function(cl) {
  set.seed(SEED)
  t0 <- Sys.time()
  res <- unc_qte(
    yname = "re78", dname = "treat", data = lalonde.psid, xformla = NULL,
    probs = QS_SIMPLE, est_method = "ipw", target = "qtt",
    cl = cl, biters = N_ITER_SIMPLE
  )
  list(
    seconds = as.numeric(Sys.time() - t0, units = "secs"),
    vals = list(
      qte = as.numeric(res$qte), se = as.numeric(res$qte.se),
      ate = as.numeric(res$ate), ate_se = as.numeric(res$ate.se)
    )
  )
}

r_simple_seq <- run_r_simple(1)
r_simple_par <- run_r_simple(N_WORKERS)
r_simple_seq_seconds <- r_simple_seq$seconds
r_simple_par_seconds <- r_simple_par$seconds
r_simple_seq_vals <- r_simple_seq$vals
r_simple_par_vals <- r_simple_par$vals
```

```{python}
#| label: py-simple
#| code-summary: "Show code (Python estimate of the unconditional QTE)"
def run_py_simple(n_workers):
    cfg = BootstrapConfig(n_iter=int(r.N_ITER_SIMPLE), seed=int(r.SEED), n_workers=n_workers)
    t0 = time.perf_counter()
    res = estimate_simple_qte(DATA, "re78", "treat", qs=qs_simple, bootstrap_config=cfg)
    return time.perf_counter() - t0, res


py_simple_seq_seconds, res_py_simple_seq = run_py_simple(1)
py_simple_par_seconds, res_py_simple_par = run_py_simple(int(r.N_WORKERS))
```

```{python}
#| label: simple-tables
#| code-summary: "Show code (Python building the simple comparison tables)"
simple_order = [f"QTE({q:.2f})" for q in qs_simple] + ["ATE"]


def py_simple_cols(res):
    rows = {
        f"QTE({row['q']:.2f})": (row["effect"], row["se"]) for row in res.qtt.iter_rows(named=True)
    }
    rows["ATE"] = (res.att["effect"][0], res.att["se"][0])
    return [rows[k][0] for k in simple_order], [rows[k][1] for k in simple_order]


def r_simple_cols(vals):
    eff = [float(x) for x in vals["qte"]] + [float(vals["ate"])]
    err = [float(x) for x in vals["se"]] + [float(vals["ate_se"])]
    return eff, err


simple_table_html = make_estimate_comparison_table(
    make_comparison_data(
        qs_simple,
        {
            ("py-qte", "Sequential"): py_simple_cols(res_py_simple_seq),
            ("py-qte", "Parallel"): py_simple_cols(res_py_simple_par),
            ("qte", "Sequential"): r_simple_cols(r.r_simple_seq_vals),
            ("qte", "Parallel"): r_simple_cols(r.r_simple_par_vals),
        },
    ),
    "Estimation under Unconfoundedness",
).as_raw_html()

data = pl.DataFrame(
    {
        "language": ["Python", "Python", "R", "R"],
        "package": ["py-qte", "py-qte", "qte", "qte"],
        "computation": ["Sequential", "Parallel", "Sequential", "Parallel"],
        "workers": [1, int(r.N_WORKERS), 1, int(r.N_WORKERS)],
        "iterations": [int(r.N_ITER_SIMPLE)] * 4,
        "seconds": [
            py_simple_seq_seconds,
            py_simple_par_seconds,
            float(r.r_simple_seq_seconds),
            float(r.r_simple_par_seconds),
        ],
    }
)

simple_timing_html = make_timing_table(data, "Estimation under Unconfoundedness").as_raw_html()
```

```{r results='asis', echo=FALSE}
cat(py$simple_table_html)
cat("<p></p>")
cat(py$simple_timing_html)
```

Notice first that first `py-qte` replicates the point estimates of the `qte` package and produces
similar standard errors. Notice that the `py-qte` produces consistent standard error estimates
across the sequential and the parallized run as it should.

We see that `py-qte` is in a _sequential scenario_ much faster than `qte`. This is probably mostly
because `np.quantile` provides an efficient implementation of computing weighted quantiles whereas
the `qte` package runs a solver for each quantile. The speed-up from parallization is very large for
`qte` and sizeable but relatively low `py-qte`.

### AIPW Estimator

```{r r-sequential}
#| code-summary: "Show code (R AIPW estimate (sequential))"
set.seed(SEED)
t0 <- Sys.time()
res_r_seq <- unc_qte(
  yname = "re78", dname = "treat", data = lalonde.psid, xformla = X_R,
  probs = QS, est_method = "aipw", target = "qte", cl = 1, biters = N_ITER
)
r_seq_seconds <- as.numeric(Sys.time() - t0, units = "secs")

r_seq_vals <- list(
  qte = as.numeric(res_r_seq$qte), se = as.numeric(res_r_seq$qte.se),
  ate = as.numeric(res_r_seq$ate), ate_se = as.numeric(res_r_seq$ate.se)
)
```

```{r r-parallel}
#| code-summary: "Show code (R AIPW estimate (parallel))"
set.seed(SEED)
t0 <- Sys.time()
res_r_par <- unc_qte(
  yname = "re78", dname = "treat", data = lalonde.psid, xformla = X_R,
  probs = QS, est_method = "aipw", target = "qte", cl = N_WORKERS, biters = N_ITER
)
r_par_seconds <- as.numeric(Sys.time() - t0, units = "secs")

r_par_vals <- list(
  qte = as.numeric(res_r_par$qte), se = as.numeric(res_r_par$qte.se),
  ate = as.numeric(res_r_par$ate), ate_se = as.numeric(res_r_par$ate.se)
)
```

```{python}
#| label: py-sequential
#| code-summary: "Show code (Python AIPW estimate (sequential))"
cfg = BootstrapConfig(n_iter=int(r.N_ITER), seed=int(r.SEED), n_workers=1)
t0 = time.perf_counter()
res_py_seq = estimate_aipw_qte(
    DATA,
    "re78",
    "treat",
    qs=qs,
    ps_x_formular=r.X_PY,
    or_x_formular=r.X_PY,
    bootstrap_config=cfg,
)
py_seq_seconds = time.perf_counter() - t0
```

```{python}
#| label: py-parallel
#| code-summary: "Show code (Python AIPW estimate (parallel))"
cfg = BootstrapConfig(n_iter=int(r.N_ITER), seed=int(r.SEED), n_workers=int(r.N_WORKERS))
t0 = time.perf_counter()
res_py_par = estimate_aipw_qte(
    DATA,
    "re78",
    "treat",
    qs=qs,
    ps_x_formular=r.X_PY,
    or_x_formular=r.X_PY,
    bootstrap_config=cfg,
)
py_par_seconds = time.perf_counter() - t0
```

```{python}
#| label: aipw-tables
#| code-summary: "Show code (Python building the AIPW comparison tables)"
order = [f"QTE({q:.2f})" for q in qs] + ["ATE"]


def py_cols(res):
    rows = {
        f"QTE({row['q']:.2f})": (row["effect"], row["se"]) for row in res.qtt.iter_rows(named=True)
    }
    rows["ATE"] = (res.att["effect"][0], res.att["se"][0])
    return [rows[k][0] for k in order], [rows[k][1] for k in order]


def r_cols(vals):
    eff = [float(x) for x in vals["qte"]] + [float(vals["ate"])]
    err = [float(x) for x in vals["se"]] + [float(vals["ate_se"])]
    return eff, err


py_seq_e, py_seq_s = py_cols(res_py_seq)
py_par_e, py_par_s = py_cols(res_py_par)
r_seq_e, r_seq_s = r_cols(r.r_seq_vals)
r_par_e, r_par_s = r_cols(r.r_par_vals)

aipw_data = make_comparison_data(
    qs,
    {
        ("py-qte", "Sequential"): (py_seq_e, py_seq_s),
        ("py-qte", "Parallel"): (py_par_e, py_par_s),
        ("qte", "Sequential"): (r_seq_e, r_seq_s),
        ("qte", "Parallel"): (r_par_e, r_par_s),
    },
)

table_html = make_estimate_comparison_table(aipw_data, "Estimation via AIPW").as_raw_html()

timing = pl.DataFrame(
    {
        "language": ["Python", "Python", "R", "R"],
        "package": ["py-qte", "py-qte", "qte", "qte"],
        "computation": ["Sequential", "Parallel", "Sequential", "Parallel"],
        "workers": [1, int(r.N_WORKERS), 1, int(r.N_WORKERS)],
        "iterations": [int(r.N_ITER)] * 4,
        "seconds": [py_seq_seconds, py_par_seconds, float(r.r_seq_seconds), float(r.r_par_seconds)],
    }
)

timing_html = make_timing_table(timing, "Estimation via AIPW").as_raw_html()
```

```{r results='asis', echo=FALSE}
cat(py$table_html)
cat("<p></p>")
cat(py$timing_html)
```

In terms of results, we see that the point estimates are close between `py-qte` and `qte`. The
standard errors are once more consistent across the sequential and parallized runs.

We see once more that the sequential version of `py-qte` is much faster than the `qte` version. This
is probably largely due to the fact that `py-qte` computes $F_{AIPW}(y)$ on the y-grid in a single
(vectorized) pass whereas the `qte` version computes each point on the y-grid separatley in a loop.
The parallized version of `py-qte` retains a sizeable advantages over the `qte` version although the
gaps is sizeably narrower than in the sequential case.

## Environment

```{r}
#| label: environment
#| code-summary: "Show code (R environment details)"
system_ <- Sys.info()[["sysname"]]


get_linux_os <- function(key) {
  if (!file.exists("/etc/os-release")) {
    return(NA_character_)
  }
  x <- grep(paste0("^", key, "="), readLines("/etc/os-release"), value = TRUE)[1]
  if (is.na(x)) {
    return(NA_character_)
  }
  value <- gsub('^"|"$', "", sub(paste0("^", "PRETTY_NAME", "="), "", x))
  if (grepl("Void Linux", value)) value <- paste0(value, " btw.")
  value
}
proc_value <- function(path, key) {
  if (!file.exists(path)) {
    return(NA_character_)
  }
  x <- grep(key, readLines(path), value = TRUE)[1]
  if (is.na(x)) {
    return(NA_character_)
  }
  trimws(sub("^[^:]*:", "", x))
}

if (system_ == "Darwin") {
  cpu_name <- system("sysctl -n machdep.cpu.brand_string", intern = TRUE)
} else {
  cpu_name <- proc_value("/proc/cpuinfo", "model name")
}

library(ps)
env_stats <- data.frame(
  item = c(
    "OS", "Kernel", "CPU", "Cores (physical)", "Cores (logical)",
    "Memory", "R", "BLAS", "Python"
  ),
  value = c(
    if (system_ == "Linux") get_linux_os() else system_,
    paste(Sys.info()[["sysname"]], Sys.info()[["release"]], Sys.info()[["machine"]]),
    cpu_name,
    parallel::detectCores(logical = FALSE),
    parallel::detectCores(logical = TRUE),
    sprintf("%.1f GiB", ps_system_memory()$total / 1e9),
    R.version.string,
    unname(extSoftVersion()["BLAS"]),
    sub(" .*", "", reticulate::py_config()$version_string)
  )
)
knitr::kable(env_stats, col.names = c("", ""), row.names = FALSE)
```
